% !TEX program = lualatex
% !TEX spellcheck = en_GB
% !TEX root = ../groups.tex

\section{Groups and subgroups}

%\begin{definition}[Groups]
A {\em group} is a pointed set \((G, 1)\) together with an {\em operation}
\[G \times G \to G \,, \ (x, y) \to x y\]
such that all this satisfies the following axioms:
\begin{enumerate}
\item \((xy)z = x(yz)\) for every \(x,y,z \in G\): this is the {\em associative property};
\item for every \(x \in G\) we have \(x1 = 1x = x\);
\item for every \(x \in G\) there is \(y \in G\), named {\em inverse} of \(x\), such that \(xy = yx = 1\).
\end{enumerate}
%\end{definition}

In most cases, there exists an obvious way for a set to give rise to a group structure.

\begin{example}
The most natural group structure upon \(\zz\) is the one that comes as you consider the usual operation of addition and \(0 \in \zz\): the addition is associative, \(0\) is the identity and for \(x \in \zz\) the element \(-x\) is the inverse of \(x\). Notice that if you replace the addition with the multiplication, the axioms (2) and (3) are violated. From now on, with \q{the group \(\zz\)}, unless otherwise specified, we mean the set \(\zz\) with \(0\) and the addition.\newline
Some numerical set (not all!) or subsets of numerical sets provide numerous examples of groups: as exercise of language, one may look for some of them.
\end{example}

\begin{example}
For a set \(X\), we have the set
\[\cals_X \coloneq \set{f : X \to X \mid f \text{ is bijective}} .\]
If you take into account the composition of functions and the identity function \(\id_X\) you will recognise a groups structure: this is the {\em symmetric group of \(X\)}! From now on, \q{the group \(\cals_X\)} is the \q{set \(\cals_X\) with \(\id_X\) and the composition}. The case when \(X\) is finite is relevant, and we adopt the following convention:
\[\cals_n \coloneq \cals_{\set{1, \dots{}, n}} \text{, where } n \in \nn^{\ge 1} .\]
\end{example}

We stick to this convention: at a purely theoretical level (definitions and theorems), we renounce to indicate the operation with a dedicated symbol (as happens in other contexts, where \(+\), \(\cdot\), \(\circ\), \dots{} are used) and simply juxtapose two elements to operate with them; unless differently said, a generic \(1\) is used to indicate the identity of a group. In that case, if \(G\) is the underlying set of a group structure, we indicate such group with the same name, \(G\), without any mention to identity and group operation.

\begin{proposition}
In any group, the identity is unique and every element has a unique inverse.
\end{proposition}

\begin{proof}
Let \(e \in G\) be an identity: \(e = e1\) because \(1\) is an identity, but also \(e1 = 1\) because \(e\) is an identity too. Thus \(e = 1\). For \(x \in G\), let \(a, b \in G\) two inverses of \(x\). We have
\[a = a 1 = a (x b) = (a x) b = 1 b = b.\qedhere\]
\end{proof}

Due to the fact identity is unique, we generically denote this element with \(1\). We write \(\inv x\) to mean the unique inverse of \(x\).

\begin{exercise}
Calculate \(\inv{(ab)}\), where \(a\) and \(b\) are elements of some group.
\end{exercise}

%As in Set Theory we there are \q{sub}sets, we want to have \q{sub}groups as well. The idea is a subgroup to be a subset that inherits the group structure from a group which contains it.

\begin{definition}
Let \(G\) be a group. A {\em subgroup} of \(G\) is a non empty set \(H\) such that
\begin{enumerate}
\item for every \(a, b \in H\) also \(a b \in H\);
\item for every \(a \in H\) also \(\inv a \in H\).
\end{enumerate}
\end{definition}

The following lemma provides a useful method to check whether a subset is a subgroup.

\begin{lemma}\label{lem:SubgroupsCond}
For \(G\) a group, any non empty \(H \subseteq G\) is a subgroup of \(G\) if and only if for every \(a, b \in H\) we have \(a \inv b \in H\).
\end{lemma}

\begin{proof}
Suppose \(H \subseteq G\) is a group. For \(b \in H\), by (2), we have \(\inv b \in H\); now, using (1), for all \(a \in H\) we have \(a\inv b \in H\). Conversely, let a non empty \(H \subseteq G\) satisfy the property: \(a \inv b \in H\) for every \(a, b \in H\). We directly have that \(\inv a \in H\) for every \(a \in H\), since \(\inv a = 1 \inv a\). Now let \(b \in H\): we have \(b = \inv{(\inv b)}\), so \(b \in H\) too; hence \(a b \in H\) for every \(a \in H\).
\end{proof}

\begin{proposition}
Consider a group \(G\), and a family of its subgroups \(\set{H_i}_{i \in I}\). Then \(\bigcap_{i \in I} H_i\) is a subgroup of \(G\). Not always \(\bigcup_{i \in I} H_i\) is a subgroup of \(G\).
\end{proposition}

\begin{proof}
The proof of the first part immediately follows form the previous Lemma. Consider \(3 \zz\) and \(5\zz\) with the operation of addition: their union is not a subgroup of \(\zz\), because for example \(8 \notin 3\zz \cup 5\zz\).
\end{proof}

\begin{exercise}
Demonstrate that the union of two subgroups is a subgroup if and only if one of them is contained by the other.
\end{exercise}

\begin{proposition}
Let \(G\) be a group and \(S \subseteq G\). There exists one and only one subgroup \(S^\ast\) of \(G\) with the following property: \(S \subseteq S^\ast\) and \(S^\ast \subseteq H\) for every subgroup \(H\) of \(G\) that contains \(S\).
\end{proposition}

\begin{proof}
Indicate with \(\mathcal I\) the family of the subgroups of \(G\) that contains \(S\). \(\mathcal I \ne \nil\) because \(G \in \mathcal I\). The subgroup \(\bigcap \mathcal I\) is what we are looking for.
\end{proof}

We write \(\gen{S}\) instead of \(S^\ast\) and we say it is the subgroup {\em generated} by \(S\). In general, those groups are quite difficult to understand and we will study the most simple case, in which \(S\) is a singleton. In that case the notation \(\gen x\) is preferred instead of \(\gen{\set{x}}\).
